Numerical simulations of the Cordilleran ice sheet through the last glacial cycle

### 2016-01-17 cordillera-cycle.tex
Typeset: more unit formatting.

      A~pseudo-plastic sliding law,
\begin{align}
&\label{eqn:pseudoplastic}
    \vec{\tau}_{\mathrm{b}} = -\tau_{\mathrm{c}} \frac{\vec{v}_{\mathrm{b}}}{{v_{\text{th}}}^q\,|\vec{v}_{\mathrm{b}}|^{1-q}} \,,
\end{align}
      relates the bed-parallel shear stresses, $\vec{\tau}_{\mathrm{b}}$, to the
      sliding velocity, $\vec{v}_{\mathrm{b}}$ (Table~\ref{tab:params}). The yield
      stress, $\tau_{\mathrm{c}}$, is modelled using the Mohr--Coulomb criterion,
\begin{align}
&\tau_{\mathrm{c}} = c_0 + N\,\tan{\phi} \,,
\end{align}
      where cohesion, $c_0$, is assumed to be zero. The friction angle,
      $\phi$, varies from 15 to 45\degree as a~piecewise-linear function of
      modern bed elevation, with the lowest value occuring below modern sea
      level (0\,\unit{m} above sea level, \unit{m\,a.s.l.}) and the highest
      value occuring above the generalised elevation of the highest
      shorelines \citep[200\,\unit{m\,a.s.l.},][Fig.~5]{Clague.1981}, thus
      accounting for a~weakening of till associated with the presence of
      marine sediments (cf. \citealp{Martin.etal.2011};
      \citealp[Supplement]{Aschwanden.etal.2013};
      \citealp{PISM-authors.2015}). Effective pressure, $N$, is related to
      the ice overburden stress, $\rho gh$, and the modelled amount of
      subglacial water, using a~formula derived from laboratory experiments
      with till extracted from the base of Ice Stream B in West Antarctica
      \citep{Tulaczyk.etal.2000, Bueler.Pelt.2015},
\begin{align}
&N = \delta \rho gh \, 10^{(e_0/C_{\mathrm{c}}) (1 - (W/W_{\text{max}}))} \,,
\end{align}
      where $\delta$ is chosen as 0.02, $e_0$ is a~measured reference void
      ratio and $C_{\mathrm{c}}$ is a~measured compressibility coefficient
      (Table~\ref{tab:params}). The amount of water at the base, $W$, varies
      from zero to $W_{\text{max}}=2$\,\unit{m}, a~threshold above which additional
      melt water is assumed to drain off instantaneously. Finally, the
      bedrock topography responds to ice load following a~bedrock
      deformation model that includes local isostasy, elastic lithosphere
      flexure and viscous astenosphere deformation in an infinite half-space
      \citep{Lingle.Clark.1985,Bueler.etal.2007}. A~relatively low viscosity
      value of $\nu_{m} = 1\times10^{19}$\,\unit{Pa\,s} is used for the
      astenosphere (Table~\ref{tab:params}) in accordance with the results
      from regional glacial isostatic adjustment modelling at the northern
      Cascadia subduction zone \citep{James.etal.2009}.

@@

      Despite large differences in the timing of attained volume extrema
      (Table~\ref{tab:extrema}), all model runs show relatively consistent
      patterns of glaciation. During MIS~4, all simulations produce an
      extensive ice sheet, covering an area of at least half of that
      attained during MIS~2 (Table~\ref{tab:extrema};
      Fig.~\ref{fig:lr_maps}, upper panels). Corresponding maximum ice
      volumes also differ significantly between model runs, and vary between
      3.84 and~8.84\,m sea level equivalents (m.~s.l.e.;
      Table~\ref{tab:extrema}).

@@

      During MIS~4, the modelled total ice volume peaks at
      7.19\,\unit{m}\,s.l.e. (57.3\,\unit{ka}) in the GRIP simulation and at
      5.04\,\unit{m}\,s.l.e. (61.9\,\unit{ka}) in the EPICA simulation,
      corresponding to respectively 90\,\% and 57\,\% of modelled MIS~2 ice
      volumes.

@@

      The modelled total ice volume corresponding to these persistent
      ice-dispersal centres attains a~minimum of 1.52\,\unit{m}\,s.l.e.
      (42.9\,\unit{ka}) in the GRIP simulation and of 2.68\,\unit{m}\,s.l.e.
      (51.8\,\unit{ka}) in the EPICA simulation, corresponding to
      respectively 19\,\% and 31\,\% of the MIS~2 ice volumes.

@@

\begin{table*}[t]
  \caption{Palaeo-temperature proxy records and scaling parameters yielding
           temperature offset time-series used to force the ice sheet model
           through the last glacial cycle (Fig.~\ref{fig:lr_ts}). $f$
           corresponds to the scaling factor adopted to yield last glacial
           maximum ice limits in the vicinity of mapped end moraines, and
           $T_{[32, 22]}$ refers to the resulting mean temperature anomaly
           during the period -32 to~-22~ka after scaling.}
  \label{tab:records}
  {\begin{tabular}{l|ccc|ccc|l}
    \tophline
    Record & Latitude & Longitude & Elev. & Proxy & $f$ & $T_{[32, 22]}$
           & Reference\\
    & & & (\unit{m\,a.s.l.}) & & & (K) & \\
    \middlehline
    GRIP     &  72{\degree}35$^{\prime}$\,N   % 72.58 (decimal)
             &  37{\degree}38$^{\prime}$\,W   % 37.64 (decimal)
             & 3238
             & \chem{\delta^{18}O}
             & 0.37 & -5.8  % -16.4126 (before scaling)
             & \citet{Dansgaard.etal.1993} \\

@@

\begin{table*}
  \caption{Extremes in Cordilleran ice sheet volume and extent corresponding to
           MIS~4, 3 and 2 for each of the six low-resolution simulations
           (Fig.~\ref{fig:lr_ts}).}
  \label{tab:extrema}
  {\begin{tabular}{l*{3}{|ccc}}
    \tophline
             & \multicolumn{3}{c}{Age (ka)}
             & \multicolumn{3}{c}{Ice extent (\unit{10^6\,km^2})}
             & \multicolumn{3}{c}{Ice volume (\unit{m}\,s.l.e.)} \\
    Record   &  MIS~4 &  MIS~3 &  MIS~2
             &  MIS~4 &  MIS~3 &  MIS~2
             &  MIS~4 &  MIS~3 &  MIS~2 \\
    \middlehline
    GRIP     &  57.45 &  42.93 &  19.14
             &   1.87 &   0.69 &   2.07
             &   7.35 &   1.69 &   8.71 \\
    NGRIP    &  60.25 &  45.86 &  22.85
             &   2.11 &   0.74 &   2.10
             &   8.84 &   1.82 &   8.76 \\
    EPICA    &  61.89 &  45.30 &  17.23
             &   1.51 &   1.01 &   2.09
             &   5.11 &   2.74 &   8.91 \\
    Vostok   &  60.86 &  41.25 &  16.84
             &   1.51 &   1.03 &   2.01
             &   5.15 &   2.88 &   8.40 \\
    ODP 1012 &  56.46 &  47.39 &  23.19
             &   1.38 &   0.90 &   2.07
             &   4.39 &   2.35 &   8.62 \\
    ODP 1020 &  60.23 &  52.97 &  20.56
             &   1.25 &   0.72 &   2.09
             &   3.84 &   1.75 &   8.81 \\
    \middlehline
    Minimum  &  61.89 &  52.97 &  23.19
             &   1.25 &   0.69 &   2.01
             &   3.84 &   1.69 &   8.40 \\
    Maximum  &  56.46 &  41.25 &  16.84
             &   2.11 &   1.03 &   2.10
             &   8.84 &   2.88 &   8.91 \\
    \bottomhline
  \end{tabular}}
  \belowtable{}
\end{table*}

@@

\begin{figure*}
  \includegraphics{lr_ts}
  \caption{Temperature offset time-series from ice core and ocean records
           (Table~\ref{tab:records}) used as palaeo-climate forcing for the ice
           sheet model (top panel), and modelled ice volume (bottom panel)
           through the last 120\,\unit{ka}. Ice volumes are expressed in meters of sea
           level equivalent (\unit{m}\,s.l.e.). Gray fields indicate Marine Oxygen
           Isotope Stage (MIS) boundaries for MIS~2 and MIS~4 according to
      a~     global compilation of benthic \chem{\delta^{18}O} records
           \citep{Lisiecki.Raymo.2005}. Hatched rectangles highlight the
           time-volume span for ice volume extremes corresponding to MIS~4
           (61.9--56.5\,\unit{ka}), MIS~3 (53.0--41.3\,\unit{ka}), and MIS~2 (LGM,
           23.2--16.8\,\unit{ka}). Dotted lines correspond to GRIP- and EPICA-driven
           5\,\unit{km}-resolution runs.}
  \label{fig:lr_ts}
\end{figure*}

@@

\begin{figure*}
  \vspace{-1.5mm}  % enough to allow the two figures to fit on one page
  \includegraphics{hr_geom_deglacage}
  \caption{Modelled age of the last deglaciation. Areas that have been covered
           only before the last glacial maximum are marked
           in green. Hatches denote re-advance of mountain-centred ice caps and
           the decaying ice sheet between 14 and 10\,\unit{ka}..
           Dashed segments \textbf{(A--D)} indicate the location of profiles used in
           Figs.~\ref{fig:hr_pf_grip} and~\ref{fig:hr_pf_epica}.}
  \label{fig:hr_geom_deglacage}
\end{figure*}

>>>>

      A~pseudo-plastic sliding law,
\begin{align}
&\label{eqn:pseudoplastic}
    \vec{\tau}_{\mathrm{b}} = -\tau_{\mathrm{c}} \frac{\vec{v}_{\mathrm{b}}}{{v_{\text{th}}}^q\,|\vec{v}_{\mathrm{b}}|^{1-q}} \,,
\end{align}
      relates the bed-parallel shear stresses, $\vec{\tau}_{\mathrm{b}}$, to the
      sliding velocity, $\vec{v}_{\mathrm{b}}$ (Table~\ref{tab:params}). The yield
      stress, $\tau_{\mathrm{c}}$, is modelled using the Mohr--Coulomb criterion,
\begin{align}
&\tau_{\mathrm{c}} = c_0 + N\,\tan{\phi} \,,
\end{align}
      where cohesion, $c_0$, is assumed to be zero. The friction angle,
      $\phi$, varies from 15 to 45\unit{\degree} as a~piecewise-linear function of
      modern bed elevation, with the lowest value occuring below modern sea
      level (0\,\unit{m} above sea level, \unit{m\,a.s.l.}) and the highest
      value occuring above the generalised elevation of the highest
      shorelines \citep[200\,\unit{m\,a.s.l.},][Fig.~5]{Clague.1981}, thus
      accounting for a~weakening of till associated with the presence of
      marine sediments (cf. \citealp{Martin.etal.2011};
      \citealp[Supplement]{Aschwanden.etal.2013};
      \citealp{PISM-authors.2015}). Effective pressure, $N$, is related to
      the ice overburden stress, $\rho gh$, and the modelled amount of
      subglacial water, using a~formula derived from laboratory experiments
      with till extracted from the base of Ice Stream B in West Antarctica
      \citep{Tulaczyk.etal.2000, Bueler.Pelt.2015},
\begin{align}
&N = \delta \rho gh \, 10^{(e_0/C_{\mathrm{c}}) (1 - (W/W_{\text{max}}))} \,,
\end{align}
      where $\delta$ is chosen as 0.02, $e_0$ is a~measured reference void
      ratio and $C_{\mathrm{c}}$ is a~measured compressibility coefficient
      (Table~\ref{tab:params}). The amount of water at the base, $W$, varies
      from zero to $W_{\text{max}}=2$\,\unit{m}, a~threshold above which additional
      melt water is assumed to drain off instantaneously. Finally, the
      bedrock topography responds to ice load following a~bedrock
      deformation model that includes local isostasy, elastic lithosphere
      flexure and viscous astenosphere deformation in an infinite half-space
      \citep{Lingle.Clark.1985,Bueler.etal.2007}. A~relatively low viscosity
      value of $\nu_{m} = 1\times10^{19}$\,\unit{Pa\,s} is used for the
      astenosphere (Table~\ref{tab:params}) in accordance with the results
      from regional glacial isostatic adjustment modelling at the northern
      Cascadia subduction zone \citep{James.etal.2009}.

@@

      Despite large differences in the timing of attained volume extrema
      (Table~\ref{tab:extrema}), all model runs show relatively consistent
      patterns of glaciation. During MIS~4, all simulations produce an
      extensive ice sheet, covering an area of at least half of that
      attained during MIS~2 (Table~\ref{tab:extrema};
      Fig.~\ref{fig:lr_maps}, upper panels). Corresponding maximum ice
      volumes also differ significantly between model runs, and vary between
      3.84 and~8.84\,\unit{m} sea level equivalents (m\,s.l.e.;
      Table~\ref{tab:extrema}).

@@

      During MIS~4, the modelled total ice volume peaks at
      7.19\,\unit{m}\,s.l.e. (57.3\,\unit{ka}) in the GRIP simulation and at
      5.04\,\unit{m}\,s.l.e. (61.9\,\unit{ka}) in the EPICA simulation,
      corresponding to respectively 90 and 57\,\unit{\%} of modelled MIS~2 ice
      volumes.

@@

      The modelled total ice volume corresponding to these persistent
      ice-dispersal centres attains a~minimum of 1.52\,\unit{m}\,s.l.e.
      (42.9\,\unit{ka}) in the GRIP simulation and of 2.68\,\unit{m}\,s.l.e.
      (51.8\,\unit{ka}) in the EPICA simulation, corresponding to
      respectively 19 and 31\,\unit{\%} of the MIS~2 ice volumes.

@@

\begin{table*}[t]
  \caption{Palaeo-temperature proxy records and scaling parameters yielding
           temperature offset time-series used to force the ice sheet model
           through the last glacial cycle (Fig.~\ref{fig:lr_ts}). $f$
           corresponds to the scaling factor adopted to yield last glacial
           maximum ice limits in the vicinity of mapped end moraines, and
           $T_{[32, 22]}$ refers to the resulting mean temperature anomaly
           during the period -32 to~-22\,\unit{ka} after scaling.}
  \label{tab:records}
  {\begin{tabular}{l|ccc|ccc|l}
    \tophline
    Record & Latitude & Longitude & Elev. & Proxy & $f$ & $T_{[32, 22]}$
           & Reference\\
    & & & (\unit{m\,a.s.l.}) & & & (K) & \\
    \middlehline
    GRIP     &  72{\degree}35$^{\prime}$\,N   % 72.58 (decimal)
             &  37{\degree}38$^{\prime}$\,W   % 37.64 (decimal)
             & 3238
             & \chem{\delta^{18}O}
             & 0.37 & -5.8  % -16.4126 (before scaling)
             & \citet{Dansgaard.etal.1993} \\

@@

\begin{table*}
  \caption{Extremes in Cordilleran ice sheet volume and extent corresponding to
           MIS~4, 3 and 2 for each of the six low-resolution simulations
           (Fig.~\ref{fig:lr_ts}).}
  \label{tab:extrema}
  {\begin{tabular}{l*{3}{|ccc}}
    \tophline
             & \multicolumn{3}{c}{Age (ka)}
             & \multicolumn{3}{c}{Ice extent (\unit{10^6\,km^2})}
             & \multicolumn{3}{c}{Ice volume (m\,s.l.e.)} \\
    Record   &  MIS~4 &  MIS~3 &  MIS~2
             &  MIS~4 &  MIS~3 &  MIS~2
             &  MIS~4 &  MIS~3 &  MIS~2 \\
    \middlehline
    GRIP     &  57.45 &  42.93 &  19.14
             &   1.87 &   0.69 &   2.07
             &   7.35 &   1.69 &   8.71 \\
    NGRIP    &  60.25 &  45.86 &  22.85
             &   2.11 &   0.74 &   2.10
             &   8.84 &   1.82 &   8.76 \\
    EPICA    &  61.89 &  45.30 &  17.23
             &   1.51 &   1.01 &   2.09
             &   5.11 &   2.74 &   8.91 \\
    Vostok   &  60.86 &  41.25 &  16.84
             &   1.51 &   1.03 &   2.01
             &   5.15 &   2.88 &   8.40 \\
    ODP 1012 &  56.46 &  47.39 &  23.19
             &   1.38 &   0.90 &   2.07
             &   4.39 &   2.35 &   8.62 \\
    ODP 1020 &  60.23 &  52.97 &  20.56
             &   1.25 &   0.72 &   2.09
             &   3.84 &   1.75 &   8.81 \\
    \middlehline
    Minimum  &  61.89 &  52.97 &  23.19
             &   1.25 &   0.69 &   2.01
             &   3.84 &   1.69 &   8.40 \\
    Maximum  &  56.46 &  41.25 &  16.84
             &   2.11 &   1.03 &   2.10
             &   8.84 &   2.88 &   8.91 \\
    \bottomhline
  \end{tabular}}
  \belowtable{}
\end{table*}

@@

\begin{figure*}
  \includegraphics{lr_ts}
  \caption{Temperature offset time-series from ice core and ocean records
           (Table~\ref{tab:records}) used as palaeo-climate forcing for the ice
           sheet model (top panel), and modelled ice volume (bottom panel)
           through the last 120\,\unit{ka}. Ice volumes are expressed in meters of sea
           level equivalent (\unit{m\,s.l.e.}). Gray fields indicate Marine Oxygen
           Isotope Stage (MIS) boundaries for MIS~2 and MIS~4 according to
           a~global compilation of benthic \chem{\delta^{18}O} records
           \citep{Lisiecki.Raymo.2005}. Hatched rectangles highlight the
           time-volume span for ice volume extremes corresponding to MIS~4
           (61.9--56.5\,\unit{ka}), MIS~3 (53.0--41.3\,\unit{ka}), and MIS~2 (LGM,
           23.2--16.8\,\unit{ka}). Dotted lines correspond to GRIP- and EPICA-driven
           5\,\unit{km}-resolution runs.}
  \label{fig:lr_ts}
\end{figure*}

@@

\begin{figure*}
  \vspace{-1.5mm}  % enough to allow the two figures to fit on one page
  \includegraphics{hr_geom_deglacage}
  \caption{Modelled age of the last deglaciation. Areas that have been covered
           only before the last glacial maximum are marked
           in green. Hatches denote re-advance of mountain-centred ice caps and
           the decaying ice sheet between 14 and 10\,\unit{ka}.
           Dashed segments \textbf{(A--D)} indicate the location of profiles used in
           Figs.~\ref{fig:hr_pf_grip} and~\ref{fig:hr_pf_epica}.}
  \label{fig:hr_geom_deglacage}
\end{figure*}
